---
title: "Wikidata SPARQL experiments"
author: "José María Mateos - http://rinzewind.org/blog-{en,es}"
date: "December 26, 2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 10
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(WikidataQueryServiceR)
```

# References

* https://www.wikidata.org/wiki/Wikidata:SPARQL_tutorial

# Questions

## When to use `wd:`, `wdt:`, `p:`, `ps:`, `pq:`?

Seems to be (rule of thumb): `p:` is a statement, `ps:` self-references a statement (see how we obtain the party in the Montreal mayor list), `pq:` is a property of a statement.

# Examples

## All Montreal mayors

```{r montreal_mayors}
query <- '
SELECT ?mayor ?mayorLabel ?start_term ?end_term ?partyLabel WHERE {
  ?mayor (wdt:P31/wdt:P279*) wd:Q5.  # look for human beings
  ?mayor wdt:P39 wd:Q177645.         # that list "Mayor of Montreal" as occupation
  ?mayor p:P39 ?statement.           # and from their term
  ?statement pq:P580 ?start_term.    # obtain the start date
  OPTIONAL {?statement pq:P582 ?end_term}.  # and the end date if present
  OPTIONAL {?mayor p:P102 [ps:P102 ?party]}. # and the party if present
  FILTER(year(?start_term) >= 1990 || year(?end_term) >= 1990) # and just take people around 1990 onwards

  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en". }
}
ORDER BY ?start_term
'

res <- query_wikidata(query)
print(res)

```

## All Madrid mayors

Should be the same as the previous one, but using `Q17078548`.

```{r madrid_mayors}
query <- '
SELECT ?mayor ?mayorLabel ?start_term ?end_term ?partyLabel WHERE {
  ?mayor (wdt:P31/wdt:P279*) wd:Q5.  # look for human beings
  ?mayor wdt:P39 wd:Q17078548.       # that list "Mayor of Madrid" as occupation
  ?mayor p:P39 ?statement.           # and from their term
  ?statement ps:P39 wd:Q17078548.    # **and only for the term as "Mayor of Madrid"**
  ?statement pq:P580 ?start_term.    # obtain the start date
  OPTIONAL {?statement pq:P582 ?end_term}.   # and the end date if present
  OPTIONAL {?mayor p:P102 [ps:P102 ?party]}. # and the party if present
  FILTER(year(?start_term) >= 1980 || year(?end_term) >= 1980) # and just take people around 1980 onwards

  SERVICE wikibase:label { bd:serviceParam wikibase:language "es". }
}
ORDER BY ?start_term
'

res <- query_wikidata(query)
print(res)
```

## Most populated countries

```{r}
query <- '
SELECT ?country ?countryLabel ?population ?area
WHERE
{
  ?country wdt:P31/wdt:P279* wd:Q6256;
           wdt:P1082 ?population;
           wdt:P2046 ?area.
  SERVICE wikibase:label {bd:serviceParam wikibase:language "en"}
}
ORDER BY DESC(?population)
LIMIT 10
'

res <- query_wikidata(query)
print(res)
```

It turns out we need to filter using actual countries. If property `P576` does not exist, that means the country currently exists.

```{r}
query <- '
SELECT DISTINCT ?country ?countryLabel ?population ?area
WHERE
{
  ?country wdt:P31/wdt:P279* wd:Q6256;
           wdt:P1082 ?population;
           wdt:P2046 ?area.
  FILTER NOT EXISTS {?country p:P576 ?x}.
  SERVICE wikibase:label {bd:serviceParam wikibase:language "en"}
}
ORDER BY DESC(?population)
LIMIT 10
'

res <- query_wikidata(query)
print(res)
```

## Replicate James Bond table

Replicate the table used in [this blog post](https://rinzewind.org/blog-en/2015/bond-movies-through-data.html).

```{r james_bond}
query <- '
SELECT ?film ?filmLabel (MIN(?publication_date) AS ?first_pub_date) ?actor_nameLabel ?box_office
WHERE {
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en". }
  # Select films of the James Bond series published in the past
  ?film (wdt:P31/wdt:P279*) wd:Q11424.      # films
  ?film wdt:P179 wd:Q2484680.               # of the James Bond series
  ?film p:P577 [ps:P577 ?publication_date]. # get the publication date
  FILTER (year(?publication_date) <= 2018)  # remove "Bond 25" from the list
  # Cast member: get who played James Bond
  ?film p:P161 ?actorStatement.             # actors in the movie
  ?actorStatement pq:P453 wd:Q2009573.      # but only the one playing Bond
  ?actorStatement ps:P161 ?actor_name.
  # Box office
  ?film p:P2142 ?box_officeStatement.
  ?box_officeStatement ps:P2142 ?box_office
}
GROUP BY ?film ?filmLabel ?actor_nameLabel ?box_office
ORDER BY ?first_pub_date
'
res <- query_wikidata(query)
print(res)
```